hbm_paper_results

### 2024-03-07 eda/ukb_curation-eda.Rmd

```{r}
ukb_df_age.filt <- ukb_df_totals %>%
  dplyr::filter(age_yrs >= 50 & age_yrs <= 80)
#write data
fwrite(ukb_df_age.filt, file=paste0(base_path,"data/ukb_CN_data_agefilt.csv"))
```

@@

```{r}
#percent female
r <- table(ukb_df_age.filt$sex)

>>>>

```{r}
ukb_df_age.filt <- ukb_df_totals %>%
  dplyr::filter(age_yrs >= 50 & age_yrs <= 80)
#write data
fwrite(ukb_df_age.filt, file=paste0(base_path,"data/ukb_CN_data_agefilt.csv"))

nrow(ukb_df_totals) - nrow(ukb_df_age.filt)
```

@@

```{r}
nrow(ukb_df_age.filt)
#percent female
r <- table(ukb_df_age.filt$sex)

### 2024-03-07 paper_lbcc_results.Rmd
getting numbers, figs

@@

@@

@@

>>>>

ComBatGAM mean centile ~ age

```{r}
gam_centile_site.mean.df <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gam_cent=mean(centile))

gam_centile_site.mean.df2 <- left_join(gam_centile_site.mean.df, delta_sum.df_merged)

gam.age.test <- lm(mean_gam_cent ~ site_mean_age, data=gam_centile_site.mean.df2)
summary(gam.age.test)
gam.age.test2 <- lm(mean_gam_cent ~ sex_prop_Female + site_n + site_mean_age, data=gam_centile_site.mean.df2)
summary(gam.age.test2)
```

NS

ComBatLS mean centile ~ age

```{r}
gamlss_centile_site.mean.df <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gamlss_cent=mean(centile))

gamlss_centile_site.mean.df2 <- left_join(gamlss_centile_site.mean.df, delta_sum.df_merged)

gamlss.age.test <- lm(mean_gamlss_cent ~ site_mean_age, data=gamlss_centile_site.mean.df2)
summary(gamlss.age.test)
gamlss.age.test2 <- lm(mean_gamlss_cent ~ sex_prop_Female + site_n + site_mean_age, data=gamlss_centile_site.mean.df2)
summary(gamlss.age.test2)
```

NS

Site-level mean centiles are not associated with study age in either ComBat-GAM or ComBatLS

@@

```{r}
cent_deltas_df %>%
  ggplot() +
  geom_histogram(aes(x=age_days)) +
  labs(title="subject distribution across age")

delta_sum.df_merged %>%
  ggplot() +
  geom_histogram(aes(x=site_mean_age_yrs)) +
  labs(title="site distribution across age")
```

@@

```{r}
ggplot(delta_sum.df_merged) +
  geom_histogram(aes(x=sex_prop_Female))
```

ComBatGAM mean centile ~ % female

```{r}
gam_centile_site.mean.df <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gam_cent=mean(centile))

gam_centile_site.mean.df2 <- left_join(gam_centile_site.mean.df, delta_sum.df_merged)

gam.prop.f.test <- lm(mean_gam_cent ~ sex_prop_Female + site_n + site_mean_age, data=gam_centile_site.mean.df2)
summary(gam.prop.f.test)
sjPlot::plot_model(gam.prop.f.test, type = "pred", terms = "sex_prop_Female")

ggplot(gam_centile_site.mean.df2, aes(x=sex_prop_Female, y=mean_gam_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```

ComBatLS mean centile ~ % female

```{r}
gamlss_centile_site.mean.df <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gamlss_cent=mean(centile))

gamlss_centile_site.mean.df2 <- left_join(gamlss_centile_site.mean.df, delta_sum.df_merged)

gamlss.prop.f.test <- lm(mean_gamlss_cent ~ sex_prop_Female + site_n + site_mean_age, data=gamlss_centile_site.mean.df2)
summary(gamlss.prop.f.test)
sjPlot::plot_model(gamlss.prop.f.test, type = "pred", terms = "sex_prop_Female")

ggplot(gamlss_centile_site.mean.df2, aes(x=sex_prop_Female, y=mean_gamlss_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
gamlss_centile_site.mean.df3 <- gamlss_centile_site.mean.df2 %>%
  mutate(combat = "gamlss") %>%
  rename(mean_cent = mean_gamlss_cent)
gam_centile_site.mean.df3 <- gam_centile_site.mean.df2 %>%
  mutate(combat = "gam") %>%
  rename(mean_cent = mean_gam_cent)

site_lvl_mean_cents <- full_join(gamlss_centile_site.mean.df3, gam_centile_site.mean.df3)

site_lvl_mean_cents %>%
  ggplot(aes(x=sex_prop_Female, y=mean_cent, color=combat, fill=combat)) +
  geom_point(alpha=0.4) +
  geom_smooth(method="lm") +
  ylab("within-site mean centile")
```

```{r}
gamlss_centile_df.mean %>%
  ggplot() + 
  geom_violin(aes(x=sex, y=mean_cent))
gamlss_centile_df.mean$combat <- "gamlss"

gam_centile_df.mean %>%
  ggplot() + 
  geom_violin(aes(x=sex, y=mean_cent))
gam_centile_df.mean$combat <- "gam"

subj.mean.cents <- full_join(gamlss_centile_df.mean, gam_centile_df.mean)

subj.mean.cents %>%
  ggplot() + 
  geom_boxplot(aes(x=sex, y=mean_cent, fill=combat))

#add sex prop
site.info <- delta_sum.df_merged %>%
  dplyr::select(site, sex_prop_Female, site_n)

subj.mean.cents2 <- left_join(subj.mean.cents, site.info)
subj.mean.cents2 %>%
  ggplot(aes(x=sex_prop_Female, y=mean_cent, color=sex)) +
  geom_point(aes(size=site_n)) +
  geom_smooth(method="lm")+
  facet_wrap(~combat)

subj.mean.cents2 %>%
  ggplot(aes(x=sex_prop_Female, y=mean_cent, color=combat, fill=combat)) +
  # geom_point(alpha=0.4) +
  geom_smooth(method="lm") +
  ylab("within-subject mean centile")
```

@@

#Exploratory: Subject-Level Mean Results

Get subject-level average deltas

```{r}
subj.avg.delta.df <- cent_deltas_df_long %>%
  mutate(abs.delta = abs(delta)) %>%
  group_by(participant) %>%
  summarise(subj_mean_abs.delta=mean(abs.delta),
            subj_mean_delta=mean(delta),
            site=unique(site),
            study=unique(study),
            sex=unique(sex))

#add back some site-level info
subj.avg.delta.sum.df <- left_join(subj.avg.delta.df, delta_sum.df_merged)

subj.avg.delta.sum.df %>%
  ggplot() +
  geom_boxplot(aes(x=sex, y=subj_mean_delta))
```

ComBat-GAM centile ~ site age

```{r}
#add back some site-level info
gam_centile_df.mean2 <- left_join(gam_centile_df.mean, delta_sum.df_merged)

summary(lm(mean_cent ~ site_mean_age, data=gam_centile_df.mean2))
```

ComBatLS centile ~ site age

```{r}
#add back some site-level info
gamlss_centile_df.mean2 <- left_join(gamlss_centile_df.mean, delta_sum.df_merged)

summary(lm(mean_cent ~ site_mean_age, data=gamlss_centile_df.mean2))
```

Neither combat method produces centiles that are assoc. with site mean age

```{r}
# age
summary(lm(subj_mean_abs.delta ~ site_mean_age, data=subj.avg.delta.sum.df))
summary(lm(subj_mean_delta ~ site_mean_age, data=subj.avg.delta.sum.df))

ggplot(subj.avg.delta.sum.df, aes(x=site_mean_age_yrs, y=subj_mean_abs.delta)) +
  geom_point() +
  geom_smooth(method="lm")
ggplot(subj.avg.delta.sum.df, aes(x=site_mean_age_yrs, y=subj_mean_delta)) +
  geom_point() +
  geom_smooth(method="lm")
```

the difference between combat and combat-ls centiles is

```{r}
#forward selection for subj_mean_abs.delta
summary(lm(subj_mean_abs.delta ~ site_mean_age_yrs + site_n, data=subj.avg.delta.sum.df))
summary(lm(subj_mean_abs.delta ~ site_mean_age_yrs + site_n + sex_prop_Female, data=subj.avg.delta.sum.df)) #small effect but remains significant/grows stronger

#forward selection for subj_mean_abs.delta
summary(lm(subj_mean_delta ~ site_mean_age_yrs + site_n, data=subj.avg.delta.sum.df))
summary(lm(subj_mean_delta ~ site_mean_age_yrs + site_n + sex_prop_Female, data=subj.avg.delta.sum.df))
```

ComBatGAM mean centile ~ % female

```{r}
gam_centile_site.mean.df <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gam_cent=mean(centile))

gam_centile_site.mean.df2 <- left_join(gam_centile_site.mean.df, delta_sum.df_merged)

gam.prop.f.test <- lm(mean_gam_cent ~ sex_prop_Female + site_n + site_mean_age, data=gam_centile_site.mean.df2)
summary(gam.prop.f.test)
sjPlot::plot_model(gam.prop.f.test, type = "pred", terms = "sex_prop_Female") 
```

ComBatLS mean centile ~ % female

```{r}
gamlss_centile_site.mean.df <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gamlss_cent=mean(centile))

gamlss_centile_site.mean.df2 <- left_join(gamlss_centile_site.mean.df, delta_sum.df_merged)

gamlss.prop.f.test <- lm(mean_gamlss_cent ~ sex_prop_Female + site_n + site_mean_age, data=gamlss_centile_site.mean.df2)
summary(gamlss.prop.f.test)
sjPlot::plot_model(gamlss.prop.f.test, type = "pred", terms = "sex_prop_Female") 
```

```{r}
gamlss_centile_df.mean %>%
  ggplot() + 
  geom_violin(aes(x=sex, y=mean_cent))
gamlss_centile_df.mean$combat <- "gamlss"

gam_centile_df.mean %>%
  ggplot() + 
  geom_violin(aes(x=sex, y=mean_cent))
gam_centile_df.mean$combat <- "gam"

subj.mean.cents <- full_join(gamlss_centile_df.mean, gam_centile_df.mean)

subj.mean.cents %>%
  ggplot() + 
  geom_boxplot(aes(x=sex, y=mean_cent, fill=combat))

#add sex prop
site.info <- delta_sum.df_merged %>%
  dplyr::select(site, sex_prop_Female, site_n)

subj.mean.cents2 <- left_join(subj.mean.cents, site.info)
subj.mean.cents2 %>%
  ggplot(aes(x=sex_prop_Female, y=mean_cent, color=sex)) +
  geom_point() +
  geom_smooth(method="lm")+
  facet_wrap(~combat)
```

```{r}
ggplot(subj.avg.delta.sum.df, aes(x=site_mean_age_yrs, y=subj_mean_abs.delta)) +
  geom_point() +
  geom_smooth(method="lm")

ggplot(subj.avg.delta.sum.df, aes(x=site_mean_age_yrs, y=subj_mean_delta)) +
  geom_point() +
  geom_smooth(method="lm")

# %female
summary(lm(subj_mean_abs.delta ~ sex_prop_Female, data=subj.avg.delta.sum.df))
summary(lm(subj_mean_delta ~ sex_prop_Female, data=subj.avg.delta.sum.df))
summary(lm(subj_mean_abs.delta ~ sex_prop_Female + site_n, data=subj.avg.delta.sum.df)) #add site size covariate
```

### 2024-03-07 paper_replication_results.Rmd

Read in centile/z-score errors (pulled centile/z error calcs from cubic using `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/ukb_permute/perm_centile_csvs/*_diffs.csv' .`)

    # Bind the data to the combined dataframe
    df_list <- c(df_list, list(data))
    assign("diffs_perm.list", df_list)
}
length(diffs_perm.list)
diffs_perm.df <- bind_rows(diffs_perm.list)
```

@@

```{r}
dataset_names <- c(ukb_CN = "Raw Data", cf = "ComBat w/o Covar", cf.lm = "ComBat", cf.gam ="ComBat-GAM", cf.gamlss ="ComBatLS")
cent.diffs.v.plot <- perm_subj.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, fill=perm), scale="count") +
  labs(y= "Mean Magnitude Centile Error", x = "ComBat Configurations") +
  scale_x_discrete(labels=c(dataset_names)) +
  guides(fill=guide_legend(title="Sampling \nPermutation")) + 
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat \nconfigurations replicated 100x") +
  theme(legend.position="none")
cent.diffs.v.plot
# ggsave(filename="figs/centile_violin.png", cent.diffs.v.plot)
```

@@

plot subj. with averages in extremes - **try also using averages w/in pheno categories**
```{r}
#recalc subject's true mean centiles
raw_mean_cent <- perm_subj.df %>%
  dplyr::filter(dataset=="cf") %>%
  mutate(mean_centile = mean_centile - mean_cent_diff,
         mean_z = mean_z - mean_z_diff) %>%
  dplyr::select(!c(Source_File, contains("diff"))) %>%
  mutate(dataset="raw")

@@

high_cent.sum <- perm_subj.df2 %>%
  dplyr::filter(mean_centile > 0.75) %>%
  group_by(dataset, perm) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

high_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  geom_hline(yintercept= high_cent.sum[high_cent.sum$dataset == "raw" & high_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

low_cent.sum <- perm_subj.df2 %>%
  dplyr::filter(mean_centile < 0.25) %>%
  group_by(dataset, perm) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

low_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  geom_hline(yintercept= low_cent.sum[low_cent.sum$dataset == "raw" & low_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed") +
  labs(title="subjects with low average centiles")

ext_cent.sum <- perm_subj.df2 %>%
  dplyr::filter(mean_centile < 0.25 | mean_centile > 0.75) %>%
  group_by(dataset, perm) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

ext_cent.sum %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=male_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  geom_hline(yintercept= ext_cent.sum[ext_cent.sum$dataset == "raw" & ext_cent.sum$perm=="perm-001",][["male_ratio"]], linetype="dashed")

@@

too many males in high centiles, too few males in low centiles (combat-GAM trends opposite to this pattern)

overall - too few men relative to women in extreme centiles

@@

>>>>

@@

```{r}
dataset_names <- c(ukb_CN = "Raw Data", cf = "ComBat w/o Covar", cf.lm = "ComBat", cf.gam ="ComBat-GAM", cf.gamlss ="ComBatLS")
cent.diffs.v.plot <- perm_subj.df %>%
  mutate(mean_cent_abs.diff = mean_cent_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_violin(aes(x=dataset, y=mean_cent_abs.diff, color=perm), scale="count") +
  labs(y= "Mean Magnitude Centile Error", x = "ComBat Configurations") +
  scale_x_discrete(labels=c(dataset_names)) +
  # guides(fill=guide_legend(title="Sampling \nPermutation")) + 
  ggtitle("Within-subject mean magnitude of error in centile scores across ComBat configurations replicated 100x") +
  theme(legend.position="none")
cent.diffs.v.plot
ggsave(filename="figs/centile_violin.png", cent.diffs.v.plot, width=11, height=4.5, units="in")
```

@@

plot subj. with averages in extremes 
```{r}
#recalc subject's true mean centiles
raw_mean_cent <- perm_subj.df %>%
  dplyr::filter(dataset=="cf") %>%
  mutate(mean_centile = mean_centile - mean_cent_diff,
        VolGlob_mean_centile = VolGlob_mean_centile - VolGlob_mean_cent_diff,
        VolReg_mean_centile = VolReg_mean_centile - VolReg_mean_cent_diff,
        CT_mean_centile = CT_mean_centile - CT_mean_cent_diff,
        SA_mean_centile = SA_mean_centile - SA_mean_cent_diff) %>%
  dplyr::select(!c(Source_File, contains("diff"))) %>%
  mutate(dataset="raw")

# raw_mean_cent %>%
#   dplyr::filter(perm=="perm-001") %>%
#   dplyr::filter(VolReg_mean_centile > .85) %>% nrow()

@@

perm_subj.df2_long <- perm_subj.df2 %>%
  rename(allpheno_mean_centile = mean_centile) %>%
  pivot_longer(cols=contains("mean_centile"), names_to="pheno_cat", values_to="mean_centile") %>%
  mutate(pheno_cat = gsub("_mean_centile", "", pheno_cat))

high_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

high_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_wrap(~pheno_cat, scales="free_y") +
  #geom_hline(yintercept=high_cent.sum[high_cent.sum$dataset == "raw" & high_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed")+
  labs(title="subjects with high average centiles")

low_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile < 0.20) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

low_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  facet_wrap(~pheno_cat, scales="free_y") +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  #geom_hline(yintercept= low_cent.sum[low_cent.sum$dataset == "raw" & low_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed") +
  labs(title="subjects with low average centiles")

ext_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile < 0.20 | mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

ext_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  #geom_hline(yintercept= ext_cent.sum[ext_cent.sum$dataset == "raw" & ext_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed")+
  facet_wrap(~pheno_cat, scales="free_y") +
  labs(title="subjects with extreme average centiles")

@@

linear ComBat -> too few women with high centiles (avg. across phenos, pattern varies), too many women with low centiles; ~ balances out overall (except regional volume)

ComBat-GAM -> too many women with high centiles (small but pretty consistent pattern), too many women with low centiles (patter varies a bit across phenos); OVERALL TOO MANY WOMEN W EXTREME CENTILES

ComBatLS -> pretty close to raw!

@@

by site
```{r}
perm_subj.df2.by.site <- perm_subj.df2 %>%
  group_by(perm, dataset, sim.site) %>%
  summarise(across(contains("mean"), list(site = mean)))

perm_subj.df2.by.site %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_cent_abs.diff_site, fill=dataset)) +
  facet_wrap(~ dataset, scales="free")

perm_subj.df2.by.site %>%
    mutate(sim.site = factor(sim.site, levels=c("Site_B", "Site_A", "Site_C"), ordered=TRUE)) %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_cent_abs.diff_site, fill=dataset)) +
  facet_wrap(~ dataset, scales="free")

perm_subj.df2.by.site %>%
  mutate(sim.site = factor(sim.site, levels=c("Site_B", "Site_A", "Site_C"), ordered=TRUE)) %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_cent_diff_site, fill=dataset)) +
  facet_wrap(~ dataset, scales="free")

perm_subj.df2.by.site %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  mutate(sim.site = factor(sim.site, levels=c("Site_B", "Site_A", "Site_C"), ordered=TRUE)) %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_cent_diff_site, fill=dataset))

perm_subj.df2.by.site %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  mutate(sim.site = factor(sim.site, levels=c("Site_B", "Site_A", "Site_C"), ordered=TRUE)) %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_centile_site, fill=dataset))

perm_subj.df2.by.site %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  pivot_wider(id_cols=c(perm, sim.site), names_from=dataset, values_from=mean_centile_site, names_prefix="site_mean_cent_") %>%
  mutate(delta_mean = site_mean_cent_cf.gam - site_mean_cent_cf.gamlss) %>%
  mutate(sim.site = factor(sim.site, levels=c("Site_B", "Site_A", "Site_C"), ordered=TRUE)) %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=delta_mean))
```
